Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-11-04 paper/paper.tex
edits to conclusion, a few edits in the middle of paper, and edits in the abstract, along with a small change to title of corr plots Tested everything on crystal.

\begin{abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust polar
  oscillations in a variety of cell shapes, new experiments have shown
  that when the cells are mechanically deformed into wide, flattened
  out, irregular shapes, the robust regularity of this oscillatory
  behavoir is no longer seen. Here we study widely used stochastic and
  deterministic models of Min-system simulation within these new cell
  shapes.  We find that the deterministic model is largely inadequate
  in reproducing the qualitatively irregular behavoir, while the
  stochastic model, based on the same differential equations, is able
  to accurately predict what has been seen.  We also find that it is
  the newly deformed cells' increased width and altered aspect ratio
  that is responsible for the irregular oscillation behavoir, as
  opposed to the irregularity of the cell shape.
\end{abstract}

@@

As described in Section~\ref{sec:model-method-shapes}, we will focus
on simulation of four illustrative flattened cell shapes: two shapes
created to replicate those shapes observed experimentally by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
cellular shape irregularity and assymetry, and that of overall size
and aspect ratio.  We note that shape \emph{B} is more irregular than
shape \emph{A}, and in particular features a prominant region of
concavity on its left-hand side.  For each of these shapes (in
addition to the wild-type pill shape discussed previously) we have
simulated in excess of 15000 seconds (over 4 hours) of evolution of
the MinD system using the deterministic version of Huang's
model~\cite{huang2003dynamic}, and in excess of 100000 seconds (over
27 hours) using the stochastic model.

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
cell shape irregularity versus overall size and aspect ratio in
developing an understanding of the origin of the experimentally
observed behavior.  Although in the movies the stadiums appear to
display, on average over long time scales, a more regular bipolar
oscillation than the irregular mannik shapes, the limited time range
of the arrow plots (350 seconds) is inadequate when studying this
difference in behavoir.  We therefore return to the correlation
function between the number of proteins in a segment at the top of the
cell and the number of proteins in a segment at the bottom of the
cell.  Fig.~\ref{fig:corr-pancake-A} shows this plot for shape
\emph{A} and stadium \emph{A}.  E. Coli cell doubling times vary
widely depending on environmental conditions, but will often be
between 20 minutes to 100 minutes~\cite{pierucci1972chromosome}, while
flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times up to 500 seconds, a biologically relevent
timescale.

@@

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.
%
The correlation function for stadium \emph{B} (5.5 periods, not shown)
is similar to that of stadium~\emph{A}, with similar coherence time,
while shape~\emph{B} exhibits a lower coherence time (2.2 periods).
In the deterministic model, shape~\emph{B} is not precisely periodic
on biologically relevant timescales, which is reflected in
Fig.~\ref{randst-plot-ave}, in which the arrows describing the
deterministic model applied to shape \emph{B} show minor maxima
forming on the right-hand side of the cell, in contrast to the regular
polar maxima formed by shape~\emph{A} and the two stadiums.
%
From the stochastic correlation data, we conclude that irregularity of
shape has only a moderate affect on the temporal regularity of the
oscillatory behavior, leaving the polar oscillation largely intact,
with even the very distorted shape~\emph{B} exhibiting only a factor
of two greater decoherence.

@@

\fixme{Stochastic model successfully explains experiment.}

\fixme{Deterministic is terrible.}
%
We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}, with a stochastic and a
deterministic version of Huang's widely used reaction-diffusion
model. We find that the deterministic model~\cite{huang2003dynamic}
predicts behavior that contradicts what is seen experimentally, namely
robust and periodic bipolar oscillation in irregularly shaped cells,
with a minor intermediate maximum appearing in shape \emph{B}.  Thus
we conclude that this method is inconsistent with experiment, and
should not be relied upon for predictions of the behavior of the MinD
system.

\fixme{What is the point here? Should it be cut?}
%
As found in previous studies~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, we observe that the stochastic model
predicts qualitative behavior in wild-type, pill-shaped cells that is
broadly consistent both with experiment and with the deterministic
model. However, even in these simple pill-shaped cells the stochastic
model shows variation in the location of MinD maxima on either end of
the cell. This model also predicts spatially irregular formation of
maxima for all the flattened cells that we studied, in agreement with
experiment.

\fixme{Observations of irregularity would be seen even in regularly
  shaped flattened or pancake cells.}
%
In order to understand the irregular MinD oscillations observed in
experiments with flattened cells, we simulated flattened ``stadium''
shapes in order to distinguish between the effect of flattening and
the effect of irregular shape.  We did not find a significant
difference in the coherence of bipolar oscillations within these
stadium shapes and the irregular shapes observed by Mannik \emph{et
  al.}.  Thus we conclude that the disruption of the MinD oscillations
seen by Mannik \emph{et al.}  is largely due to the flattened cell
shape and greater than average size, as opposed to its irregularity.

\fixme{Talk about flattened cells throughout vs. irregular cells.}

>>>>

\begin{abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust polar
  oscillations in a variety of cell shapes, new experiments have shown
  that when the cells are mechanically deformed into wide, flattened
  out, irregular shapes, the robust regularity of this oscillatory
  behavoir is no longer seen. Here we study widely used stochastic and
  deterministic models of Min-system simulation within these new cell
  shapes.  We find that the deterministic model is largely inadequate
  in reproducing the qualitatively irregular behavoir, while the
  stochastic model, based on the same differential equations, is able
  to accurately predict what has been seen.  We also find that it is
  the flattened nature of the newly deformed cells' shape that is
  responsible for the irregular oscillation behavoir, as opposed to
  its irregularity and asymmetry.
\end{abstract}

@@

As described in Section~\ref{sec:model-method-shapes}, we will focus
on simulation of four illustrative flattened cell shapes: two shapes
created to replicate those shapes observed experimentally by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
flattening the cell and the cell shape's irregularity and asymmetry.
We note that shape \emph{B} is more irregular than shape \emph{A}, and
in particular features a prominant region of concavity on its
left-hand side.  For each of these shapes (in addition to the
wild-type pill shape discussed previously) we have simulated in excess
of 15000 seconds (over 4 hours) of evolution of the MinD system using
the deterministic version of Huang's model~\cite{huang2003dynamic},
and in excess of 100000 seconds (over 27 hours) using the stochastic
model.

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening of the cell versus cell shape irregularity and
asymmetry in developing an understanding of the origin of the
experimentally observed behavior.  Although in the movies the stadiums
appear to display, on average over long time scales, a more regular
bipolar oscillation than the irregular mannik shapes, the limited time
range of the arrow plots (350 seconds) is inadequate when studying
this difference in behavoir.  We therefore return to the correlation
function between the number of proteins in a segment at the top of the
cell and the number of proteins in a segment at the bottom of the
cell.  Fig.~\ref{fig:corr-pancake-A} shows this plot for shape
\emph{A} and stadium \emph{A}.  E. Coli cell doubling times vary
widely depending on environmental conditions, but will often be
between 20 minutes to 100 minutes~\cite{pierucci1972chromosome}, while
flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times up to 500 seconds, a biologically relevent
timescale.

@@

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.
%
The correlation function for stadium \emph{B} is similar to that of
stadium~\emph{A}, with similar coherence time (5.5 periods, not
shown), while shape~\emph{B} exhibits a lower coherence time (2.2
periods).  In the deterministic model, shape~\emph{B} is not precisely
periodic on biologically relevant timescales, which is reflected in
Fig.~\ref{randst-plot-ave}, in which the arrows describing the
deterministic model applied to shape \emph{B} show minor maxima
forming on the right-hand side of the cell, in contrast to the regular
polar maxima formed by shape~\emph{A} and the two stadiums.
%
From the stochastic correlation data, we conclude that irregularity of
shape has only a moderate affect on the temporal regularity of the
oscillatory behavior, with even the very distorted shape~\emph{B}
exhibiting only a factor of two greater decoherence.

@@

We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}, with a stochastic and a
deterministic version of Huang's widely used reaction-diffusion model.
We find that the stochastic model predicts experimentally accurate
qualitative behavoir in both the wild-type, pill-shaped cells (as
found in previous studies~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}), as well as in the new, flattened,
irregularly shaped cells.  These flattened cells invariably show
spatially irregular formation of maxima that is also seen
experimentally.

We find that the deterministic model~\cite{huang2003dynamic}
predicts behavior that contradicts what is seen experimentally, namely
robust and periodic bipolar oscillation in irregularly shaped cells,
with a minor intermediate maximum appearing in shape \emph{B}.  Thus
we conclude that this method is inconsistent with experiment, and
should not be relied upon for predictions of the behavior of the MinD
system.

In order to understand the irregular MinD oscillations observed in
experiments with flattened cells, we simulated flattened ``stadium''
shapes in order to distinguish between the effect of flattening and
the effect of irregular shape.  We found only a moderate difference in
the coherence of bipolar oscillations within these stadium shapes and
the irregular shapes observed by Mannik \emph{et al.}, and have
observed visually the same qualitative spatially irregular formation
of maxima in both.  Thus we conclude that the disruption of the MinD
oscillations seen by Mannik \emph{et al.}  is largely due to the
flattening of the cell shape, as opposed to its irregularity.
